\pkg{Rgbp}: An \proglang{R} Package for Conjugate Gaussian, Poisson, and Binomial Hierarchical Modeling and Frequency Method Checking on Overdispersed Data

### 2015-05-05 jss/article.tex
introduction and abstract

%% almost as usual
\author{Hyungsuk Tak\\Harvard University \And 
             Joseph Kelly\\Google\And
             Carl Morris\\ Harvard University}
\title{\pkg{Rgbp}: An \proglang{R} Package for Gaussian, Poisson, and Binomial Hierarchical Modeling}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Joseph Kelly, Carl Morris, Hyungsuk Tak} %% comma-separated
\Plaintitle{Rgbp: An R Package for Hierarchical modeling and Method Checking} %% without formatting

%% an abstract and keywords
\Abstract{
\pkg{Rgbp} is an \proglang{R} package that utilizes approximate Bayesian machinery to provide a method of estimating two-level hierarchical models for over-dispersed Gaussian, Poisson, and Binomial data in a fast and computationally efficient manner. The main products of this package are point and interval estimates for each group's random effect, $\mu_{j}$ for Gaussian, $\lambda_{j}$ for Poisson, or $p_{j}$ for Binomial model, whose good frequency properties can be validated via its repeated sampling procedure called ``frequency method checking."  It is found that such Bayesian-frequentist reconciliation allows \pkg{Rgbp} to have attributes desirable from both perspectives, working well in small samples and yielding good coverage probabilities for its interval estimates.}
\Keywords{multilevel model, conjugate hierarchical generalized linear models, frequency method checking, coverage probability, shrinkage, \proglang{R}}
\Plainkeywords{multilevel model, conjugate hierarchical generalized linear models, frequency method checking, coverage probability, shrinkage, R} 
%% without formatting
%% at least one keyword must be supplied

>>>>

%% almost as usual
\author{Hyungsuk Tak\\Harvard University \And 
             Joseph Kelly\\Google\And
             Carl Morris\\ Harvard University}
\title{\pkg{Rgbp}: An \proglang{R} Package for Gaussian, Binomial, and Poisson Hierarchical Modeling}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Hyungsuk Tak, Joseph Kelly, Carl Morris} %% comma-separated
\Plaintitle{Rgbp: An R Package for Hierarchical modeling and Frequency Method Checking} %% without formatting

%% an abstract and keywords
\Abstract{
\pkg{Rgbp} is an \proglang{R} package that utilizes  approximate Bayesian machinery to provide a method of estimating two-level hierarchical models for over-dispersed Gaussian, Binomial, and Poisson data in a fast and computationally efficient manner. The main products of this package are point and interval estimates for each group's random effect, $\mu_j$ for Gaussian, $\lambda_j$ for Poisson, or $p_j$ for Binomial model, whose good frequency properties can be validated via its repeated sampling procedure called ``frequency method checking.'' It is found that such Bayesian-frequentist reconciliation allows \pkg{Rgbp} to have attributes desirable from both perspectives, working well in small samples and yielding good coverage probabilities for its interval estimates. }
%It is found that such Bayesian-frequentist reconciliation allows \pkg{Rgbp} to have attributes desirable from both perspectives, working well in small samples and yielding good coverage probabilities for its interval estimates.
\Keywords{overdispersion, hierarchical model, adjustment for density maximization, repeated sampling coverage evaluation, \proglang{R}}
\Plainkeywords{multilevel model, conjugate hierarchical generalized linear models, frequency method checking, coverage probability, shrinkage, R} 
%% without formatting
%% at least one keyword must be supplied

### 2015-05-05 jss/article_v2.tex
introduction and abstract

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Hyungsuk Tak, Joseph Kelly, Carl Morris} %% comma-separated
\Plaintitle{Rgbp: Hierarchical Modeling and Frequency Method Checking} %% without formatting
\Shorttitle{\pkg{Rgbp}: Hierarchical Modeling and Frequency Method Checking on over dispersed data} %% a short title (if necessary)
%% an abstract and keywords
\Abstract{
\pkg{Rgbp} is an \proglang{R} package that utilizes  approximate Bayesian machinery to fit two-level conjugate hierarchical models on overdispersed Gaussian, Poisson, and Binomial data and evaluates whether the resulting approximate Bayesian interval estimates for random effects meet the nominal confidence levels via frequency coverage evaluation. The data that \pkg{Rgbp} assumes comprise observed sufficient statistics for each random effect, such as an average or a proportion, without population-level data.  The approximate Bayesian tool equipped with the adjustment for density maximization produces approximate point and interval estimates for model parameters including second-level variance component, regression coefficients, and random effects. For the Binomial data, the package provides an option to produce  posterior samples of all the model parameters via the acceptance-rejection method. The package provides a quick way to evaluate coverage rates of the resultant Bayesian interval estimates for random effects via a parametric bootstrapping, which we call  ``frequency method checking."  }
%It is found that such Bayesian-frequentist reconciliation allows \pkg{Rgbp} to have attributes desirable from both perspectives, working well in small samples and yielding good coverage probabilities for its interval estimates.
\Keywords{overdispersion, hierarchical model, adjustment for density maximization, repeated sampling coverage evaluation, \proglang{R}}
\Plainkeywords{multilevel model, conjugate hierarchical generalized linear models, frequency method checking, coverage probability, shrinkage, R} 
%% without formatting
%% at least one keyword must be supplied

@@

\section[Introduction]{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.
Gaussian, Poisson, or Binomial data from several independent groups sometimes have more variation than the assumed Gaussian, Poisson, or Binomial distributions on the first-level observed data. To account for the extra-variability, called overdispersion, a two-level hierarchical model regards first-level mean parameters as random effects that come from a population-level conjugate prior distribution. The conjugate prior distribution is exchangeable if no covariates are available and non-exchangeable if the model incorporates covariate information of each group via a linear, log-linear, or logistic regression according to the data type.

With an assumption of homogeneity within each group, the observed data are sufficient statistics for the random effects, such as averages or proportions, possibly together with each group's covariate information. This type of  data is common for a biological analysis on litter data, a meta analysis, or small area estimation problems.  For these data, the two-level model that \pkg{Rgbp} assumes can be considered as a conjugate hierarchical generalized linear model \citep{hglm2006, hglm2010} without any fixed effects.

\pkg{Rgbp} takes a Bayesian approach with our special improper hyper-prior distributions on hyper-parameters, the parameters of the conjugate prior distribution. The hyper-prior distributions lead to Stein's harmonic prior known to produce good repeated sampling coverage properties of the Bayesian interval estimates for random effects in a two-level Gaussian hierarchical model \citep{tang2011, morris2012}.  We apply the hyper-prior distributions to Poisson and Binomial hierarchical models. The goal of our hierarchical modeling is to compare random effects among groups by their Bayesian point and interval estimates that have good frequency properties.

When it comes to fitting the model, \pkg{Rgbp} adopts the adjustment for density maximization \citep{carl1988, 1997, tang2011} (ADM), a generalization of a commonly-used restricted maximum likelihood method (REML). The ADM uses Beta distributions to approximate the posterior distributions of shrinkage factors, functions of a second-level variance component, via maximization like a Delta method that uses a Normal distribution to obtain the approximate distribution of a function of a parameter with its MLE and observed information plugged-in.  The ADM is more appropriate than the Delta method (or REML) because the support of each shrinkage factor is between 0 and 1, not ($-\infty,~\infty$). Also, the ADM is free of boundary effects that REML suffers from (e.g. a zero estimate for the second-level variance component). Using the approximate estimates for the posterior mean and variance of the shrinkage factors, \pkg{Rgbp} obtains  approximate posterior mean and variance of the random effect. Finally, \pkg{Rgbp} approximates the posterior distributions of random effects by a Pearson family distribution whose two parameters are matched to the estimated posterior mean and variance of random effects.

For the Gaussian hierarchical model, \pkg{Rgbp} approximates the posterior distribution of the random effects by a skewed Normal distribution whose three parameters are matched to the first three posterior moment estimates for random effects. For the Binomial  model, \pkg{Rgbp} provides an option to draw independent posterior samples of all the model parameters including second-level variance component, logistic regression coefficients, and random effects,  via an acceptance-rejection method instead of the approximate tool.

In addition to fitting hierarchical models, \pkg{Rgbp} provides a quick way to evaluate the repeated sampling coverage rates of the resulting Bayesian interval estimates for random effects. It is a unique procedure that distinguishes \pkg{Rgbp}  from any other \proglang{R} packages for similar multilevel models, such as \pkg{hglm} for conjugate hierarchical generalized models, \pkg{arm} for Bayesian hierarchical models, and \pkg{hbsae} for hierarchical Bayesian small area estimation. The evaluation procedure which we call ``frequency method checking'' adopts a parametric bootstrapping that generates mock data sets given hyper-parameter values and estimates the coverage rates.

%This package is useful for both frequentist and Bayesian. Bayesians are able to use the package to see results for a non-informative reference prior before and after constructing their own Bayesian hierarchical model and frequentists are provided with a procedure with good repeated sampling properties.

\section[Hierarchical Structure]{Hierarchical structure} \label{hierarchical}
A two-level or multilevel model is a powerful tool for exploring the hierarchical structure in data. For example, we can imagine that there exists a district-level hierarchy (bigger population) for observed school-level data, or a state-level hierarchy for observed hospital-level data in a certain state.

@@

\section[Estimation]{Estimation}

@@

\section[Frequency method checking]{Frequency Method Checking}
Like the two sides of the same coin, checking a statistical model comes hand in hand with fitting the model. If a fitted model cannot pass a validation or checking process, we usually go back and forth from estimation and checking steps iteratively. In this sense, checking a fitted model is an interactive procedure for the model justification.

@@

\section[Examples]{Examples}
\subsection[Known Second-level Mean]{Data of 31 hospitals with a known second-level mean}
\label{sec:ex:hosp}
In this example we adopt the perspective of a person living in the state of New York (NY) who has been suffering from severe coronary heart disease. If this person must receive coronary artery bypass graft (CABG) surgery soon, he or she might want to find the most reliable hospital for such a procedure.

>>>>

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Hyungsuk Tak, Joseph Kelly, Carl Morris} %% comma-separated
\Plaintitle{Rgbp: Hierarchical Modeling and Frequency Method Checking} %% without formatting
\Shorttitle{\pkg{Rgbp}: Hierarchical Modeling and Frequency Method Checking on over dispersed data} %% a short title (if necessary)
%% an abstract and keywords
\Abstract{
\pkg{Rgbp} is an \proglang{R} package that utilizes  approximate Bayesian machinery to fit two-level conjugate hierarchical models on overdispersed Gaussian, Poisson, and Binomial data and evaluates repeated sampling frequency property of the resulting approximate Bayesian interval estimates for random effects. The main goal of \pkg{Rgb} is to produce Bayesian interval estimates for the random effects that meet the nominal confidence level, and the package uses special improper hyper-prior distributions for that purpose. The data that \pkg{Rgbp} assumes comprise observed sufficient statistics for each random effect, such as averages or proportions, without population-level data.  The approximate Bayesian tool equipped with the adjustment for density maximization produces approximate point and interval estimates for model parameters including second-level variance component, regression coefficients, and random effects. For the Binomial data, the package provides an option to produce  posterior samples of all the model parameters via the acceptance-rejection method. The package provides a quick way to check whether the resultant Bayesian interval estimates for the random effects meet the nominal confidence levels via a repeated sampling coverage evaluation, which we call  ``frequency method checking."  }
%It is found that such Bayesian-frequentist reconciliation allows \pkg{Rgbp} to have attributes desirable from both perspectives, working well in small samples and yielding good coverage probabilities for its interval estimates.
\Keywords{overdispersion, hierarchical model, adjustment for density maximization, repeated sampling coverage evaluation, \proglang{R}}
\Plainkeywords{multilevel model, conjugate hierarchical generalized linear models, frequency method checking, coverage probability, shrinkage, R} 
%% without formatting
%% at least one keyword must be supplied

@@

\section[Introduction]{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.
Gaussian, Poisson, or Binomial data from several independent groups sometimes have more variation than the assumed Gaussian, Poisson, or Binomial distributions of the first-level observed data. To account for the extra-variability, called overdispersion, a two-level hierarchical model regards first-level mean parameters as random effects that come from a population-level conjugate prior distribution. The conjugate prior distribution is exchangeable if no covariates are available and non-exchangeable if the model incorporates covariate information of each group via a linear, log-linear, or logistic regression according to the data type.

With an assumption of homogeneity within each group, the observed data are sufficient statistics for the random effects, such as averages or proportions, possibly together with each group's covariate information. This type of  data is common for a biological analysis on litter data, a meta analysis on independent studies, or small area estimation problems.  For these data, the two-level model that \pkg{Rgbp} assumes can be considered as a conjugate hierarchical generalized linear model \citep{lee1996hierarchical, hglm2006} where each random effect has a conjugate prior distribution.

\pkg{Rgbp} takes a Bayesian approach with our special improper hyper-prior distributions on hyper-parameters, the parameters of the conjugate prior distribution, to produce  Bayesian interval   estimates for the random effects that achieve nominal confidence levels. The hyper-prior distributions lead to Stein's harmonic prior known to produce good repeated sampling coverage properties of the Bayesian interval estimates for random effects in a two-level Gaussian hierarchical model \citep{tang2011, morris2012}.  We apply the hyper-prior distributions to Poisson and Binomial hierarchical models. 
%The main goal of our Bayesian hierarchical models is to compare random effects between groups by their Bayesian interval estimates that have good frequency properties.

When it comes to fitting the model, \pkg{Rgbp} adopts the adjustment for density maximization \citep{carl1988, morris1997, tang2011} (ADM), a Pearson family approximation via maximization. For example, a Delta method is a special case of ADM that uses a Normal distribution to obtain an approximate distribution of a function of a parameter with its MLE and observed information plugged-in. In this article, the ADM uses Beta distributions to approximate the posterior distributions of shrinkage factors, functions of the second-level variance component, via maximization. For this approximation, the ADM  is more appropriate than the Delta method because the support of each shrinkage factor is between 0 and 1, not ($-\infty,~\infty$). Also, the ADM is free of a boundary effect that leads to a zero estimate for the second-level variance component  which a restricted MLE \citep{patterson1971recovery} (REML) sometimes suffers from. Using the approximate posterior distributions of shrinkage factors, \pkg{Rgbp} estimates the first two (three for the Gaussian case) posterior moments of the random effects. Finally, \pkg{Rgbp} approximates the posterior distribution of each random effect by a skewed Normal distribution for a Gaussian case, a Gamma distribution for a Poisson case, and a Beta distribution for a Binomial case whose two (three for the Gaussian case) parameters are matched to the estimated posterior moments of the random effects.

For the Binomial  model, \pkg{Rgbp} provides an option to draw independent posterior samples of all the model parameters via an acceptance-rejection method instead of the approximate tool.

In addition to fitting hierarchical models, \pkg{Rgbp} provides a quick way to evaluate the repeated sampling coverage rates of the resulting Bayesian interval estimates for random effects \citep{morris1997, daniels1999prior, tang2002fitting, tang2011, morris2012}. It is a unique procedure that distinguishes \pkg{Rgbp}  from any other \proglang{R} packages  such as \pkg{hglm} \citep{hglm2010, ronnegaard2011hglm} for conjugate hierarchical generalized models and \pkg{arm} \citep{gelman2014arm} for Bayesian hierarchical models with proper hyper-prior distributions. The evaluation procedure which we call ``frequency method checking'' adopts a parametric bootstrapping that generates mock data sets given the values of hyper-parameters and estimates the coverage rates.

The rest of this paper is organized as follows. We specify the Bayesian hierarchical models and discuss their posterior propriety in Section \ref{sec2} We describe the estimation procedures including the ADM and the acceptance-rejection method in Section  \ref{sec3}, and the frequency method checking in Section \ref{sec4}.  We explain the usages of main functions in \pkg{Rgbp} in Section \ref{sec5}, and apply them to three examples in Section \ref{sec6}.

\section[Hierarchical Structure]{Hierarchical structure} \label{sec2}
A two-level or multilevel model is a powerful tool for exploring the hierarchical structure in data. For example, we can imagine that there exists a district-level hierarchy (bigger population) for observed school-level data, or a state-level hierarchy for observed hospital-level data in a certain state.

@@

\section[Estimation]{Eestimation}\label{sec3}

@@

\section[Frequency method checking]{Frequency Method Checking}\label{sec4}
Like the two sides of the same coin, checking a statistical model comes hand in hand with fitting the model. If a fitted model cannot pass a validation or checking process, we usually go back and forth from estimation and checking steps iteratively. In this sense, checking a fitted model is an interactive procedure for the model justification.

@@

\section[packageoverview]{Main Functions in \pkg{Rgbp}}\label{sec5}

\section[Examples]{Examples}\label{sec6}
\subsection[Known Second-level Mean]{Data of 31 hospitals with a known second-level mean}
\label{sec:ex:hosp}
In this example we adopt the perspective of a person living in the state of New York (NY) who has been suffering from severe coronary heart disease. If this person must receive coronary artery bypass graft (CABG) surgery soon, he or she might want to find the most reliable hospital for such a procedure.
